Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex
spurious 'the'
typos

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space. The notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
tests the consistency of a modal $M_C$ of source code against
a model $M_S$ of a specification:
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
function to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by the \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite thus equality cannot be computed by exhaustively
comparing elements in each set. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
query is expressed in quantifier free linear arithmetic, which is decidable.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are
connected with no more than two \texttt{+} operations in 100\% of the cases.

\subsection{Inferring specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code which are inserted automatically as comments. This supports
maintenance of legacy code base, and aids understanding of our
specification language.

The preceding source code analysis converts the concrete syntax of
array subscripts into a sets of subscript schemes. Inference then has
two parts. First, ``adjacent'' index schemes are coalesced into a
smaller sets of index schemes, remaining in union normal
form. Secondly, the resulting union normal form is converted to the
specification syntax.

@@

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing scheme $\mathcal{S}$ and a particular index scheme $S$
we generate the set of coalescing of $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This give a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$, then
  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

\newcommand{\finalSet}{\mathcal{S}_\omega}
Let the covering indexing scheme from the fixed point of
$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
$\finalSet{}$.  Next, $\finalSet{}$ the is translated into specification
syntax. This happens in four stages.

@@

If there is all infinite indexing scheme, then we cannot represent it
syntactically. This happens when use of constant indices in one or more array
terms causes the offsets to be subsumed. The simplest case is when there is an
array term with all of its indices constant.

@@

After handling the trivial case, correct approximation is chosen. Recall
\Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
the sets in products are holed intervals. If that is the case, we trivially
convert the sets in each dimension of indexing schemes to holed interval
representation and proceed to the next stage and output an exact specification
at the end. If not, then the output will definitely have an upper bound and
potentially a lower bound.

Upper bound is established by \emph{elongating} index schemes in multiple
dimensions such that they become interval schemes. The elongation function is
given as
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}

Since index schemes of the previous stage are all contiguous and form closed
intervals, 0 lying between the lower and upper bounds of intervals is sufficient
to form an interval scheme. If any dimension of an index scheme is elongated,
then the whole index scheme is elongated. In the initial set of index schemes,
all schemes that were already interval schemes form a lower bound, while those
schemes and the new interval schemes generated by elongation form the upper
bound. The reason a lower bound may not be generated is all indexing schemes
with might need elongation.

@@

There is no direct translation from an interval scheme such as $\interv{-2}{1}{}
\times \interv{-1}{1}{}$ to region constants. The reason is region constants
\texttt{forward} and \texttt{backward} require upper and lower bounds to be 0
respectively and \texttt{centered} requires the upper bound to be the negation
of the lower bound. The solution is to use \Cref{lem:vector-union} to break up
interval schemes into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
and $\interv{0}{1}{} \times \interv{-1}{1}{}$.

@@

In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. Decomposition of
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}

>>>>

\noindent
Checking verifies the access pattern of an array computation in the
source language against any associated specifications. Checking
proceeds by generating a model from a specification and generating a
model from the source code (above), and comparing them for
consistency.  Since the model of \Cref{sec:model} interprets both
array indices and specifications as sets of points in $N$-dimensional
space. The notion of consistency is then intuitively whether these two
(potentially infinite) sets of points are equal. This leads to a
simple notion of consistency, where $\mathit{consistent}(M_C, M_S)$
tests the consistency of a modal $M_C$ of source code against
a model $M_S$ of a specification:
%
\begin{align*}
  \mathit{consistent} & :
    \mathsf{Mult}(\mathcal{P}(\bz{}^N)) \times
    \mathsf{Mult}(\mathsf{Approx}(\region{N})) \to \mathbb{B} & \\
%
  \mathit{consistent}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    \mathsf{false} & \mathit{model} = \mathsf{once}(x) \wedge ix =
    \mathsf{mult}(y) \\
    \mathsf{false} & \mathit{model} = \mathsf{mult}(x) \wedge ix = \mathsf{once}(y) \\
    \mathit{consistent'}(\mathit{peel}(ix), \mathit{peel}(model)) & \textit{otherwise}
  \end{cases} \\
%
  \mathit{consistent'} & :
    \mathcal{P}(\bz{}^N) \times
    \mathsf{Approx}(\region{N}) \to \mathbb{B} & \\
%
  \mathit{consistent'}(\mathit{ix}, \mathit{model}) & = \begin{cases}
    m = \mathit{ix} & \mathit{model} = \mathsf{exact}(m) \\
    m \supseteq \mathit{ix} & \mathit{model} = \mathsf{upper}(m) \\
    m \subseteq \mathit{ix} & \mathit{model} = \mathsf{lower}(m)
  \end{cases}
\end{align*}
%
The first function, $\mathit{consistent}$, checks whether the linearity of the
specification matches that of the indices, \ie{} if the specification allows
indices to be repeated or not. Then it delegates to $\mathit{consistent'}$
to check if the points observed in the array terms match the space
defined by the specification. Lower bounds, marked by the \texttt{atLeast}
modifier, require the space defined by the specification to remain inside the
set of indices, while an upper bound, market by \texttt{atMost},
requires the opposite: enclosure. In the absence of such modifiers, we expect
the space defined by observed indices corresponds exactly with those defined by the
region in the specification, hence requiring set equality.

As explained in \Cref{subsec:union-normal-form}, the sets being compared are
potentially infinite thus equality cannot be computed by exhaustively
comparing elements in each set. Instead, we compile the region into
interval constraints and subscript schemes into membership constraints and pass
these to the \textsc{Z3} SMT solver~\citep{de2008z3} to see if they are equal. The
query is expressed in quantifier-free linear arithmetic, which is decidable.

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, the performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times the number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are
connected with no more than two \texttt{+} operations in 100\% of the cases.

\subsection{Inferring specification automatically}
\label{subsec:inference}
%
\noindent
We provide an inference procedure for generating specifications from
code which are inserted automatically as comments. This supports
maintenance of a legacy code base, and aids adoption of the
specification language.

The preceding source code analysis converts the concrete syntax of
array subscripts into sets of subscript schemes. Inference then has
two parts. First, ``adjacent'' index schemes are coalesced into a
smaller sets of index schemes, remaining in union normal
form. Secondly, the resulting union normal form is converted to the
specification syntax.

@@

\newcommand{\contig}[2]{\mathit{adjacent}(#1, #2)}
\begin{defn}[Adjacent]
  \label{def:contiguity}
  Two index schemes $S$ and $T$ are \emph{adjacent} written
  $\contig{S}{T}$, iff $\pi_i(S) = \pi_i(T)$ for all $1 \leq i \leq N$
  apart from some dimension $k$ such that $\pi_k(S) = [a, b]$ and
  $\pi_k(T) = [b+1, c]$.
\end{defn}
%
Given a set of indexing schemes $\mathcal{S}$ and a particular index scheme $S$
we generate a set from coalescing $S$ with adjacent index schemes:
%
\begin{equation*}
\mathit{coalesce}(S, \mathcal{S}) = 
  \{\,S \cup T \mid T \in \mathcal{S} \, \wedge \,
  \contig{S}{T} \}
\end{equation*}
%
This is then used by the following recursive procedure:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}) =
  \exists T \in \mathcal{S} . 
  \begin{cases}
    \{T\} \cup \mathit{coalesceStep}(\mathcal{S} - \{T\}) &
    \mathit{coalesce}(T, \mathcal{S}) = \emptyset \\
    \mathit{coalesceStep}(\mathit{coalesce}(T, \mathcal{S}) \cup
                                        \mathcal{S} - \{T\}) & \textit{otherwise}
  \end{cases}
\end{align*}
%
This gives a specification rather than an implementation. Our
implementation represents sets by a list, and so
$\exists T \in \mathcal{S}$ above corresponds to deconstructing
the list into its head element $S$ (picking an element from the set).
If $T$ has no adjacent index schemes, then coalescing is attempted on
the rest of the set, and $T$ is returned in the result. Otherwise,
the set of coalesced index scheme is computed and this is passed to
a recursive called to $\mathit{coalesceStep}$, along with the rest of
the elements).

The fixed-point of $\textit{coalesceStep}(\mathcal{S})$ is computed to
give a covering over the initial subscript space.
%
\begin{restatable}{lem}{closureInference}
  \label{lem:closuer-inference}
  For a set of index schemes $\mathcal{S}$,
  $\mathit{coalesceSet}(\mathcal{S})$ is well-defined and is a set of
  index schemes in Union Normal Form.
\end{restatable}
%
For our example, the fixed-pointed of $\mathit{coalesceStep}$ is reached within two
steps:
%
\begin{align*}
  \mathit{coalesceStep}(\mathcal{S}_0) & =
  \{[-1,0] \times [0,0],\,[0,0] \times [-1, 0],\,[0,0] \times
    [0,1],\,[0,1] \times [0,0]\} \\
  \mathit{coalesceStep}^2(\mathcal{S}_0) & =
  \{[-1, 1] \times [0, 0],\,[0, 0] \times [-1, 1]\} = \mathit{coalesceStep}^3(\mathcal{S}_0)
\end{align*}
%
%\todo{Prove that this is in $Region_N$}

@@

\newcommand{\finalSet}{\mathcal{S}_\omega}
Let the covering indexing scheme from the fixed point of
$\mathit{coalesceStep}$ on input $\mathcal{S}_0$ be written as
$\finalSet{}$.  Next, $\finalSet{}$ is translated into specification
syntax. This happens in four stages.

@@

If the result is an all infinite indexing scheme, then we cannot represent it
syntactically. This happens when use of constant indices in one or more array
terms causes the offsets to be subsumed. The simplest case is when there is an
array term with all of its indices constant.

@@

After handling the trivial case, the correct approximation is chosen. Recall
\Cref{def:interval-scheme}, an indexing scheme is an interval scheme if each of
the sets in products are holed intervals. If that is the case, we trivially
convert the sets in each dimension of the indexing schemes to a holed interval
representation and proceed to the next stage and output an exact specification
at the end. If not, then the output will definitely have an upper bound and
potentially a lower bound.

The upper bound is established by \emph{elongating} index schemes in multiple
dimensions such that they become interval schemes. The elongation function is
given as
%
\begin{align*}
  \mathit{elongate}([a,b]) = \begin{cases}
    \interv{0}{b}{\textsf{false}} & a > 1 \\
    \interv{a}{0}{\textsf{false}} & b < -1
  \end{cases}
\end{align*}

Since the index schemes from the previous stage are all contiguous and form closed
intervals, 0 lying between the lower and upper bounds of intervals is sufficient
to form an interval scheme. If any dimension of an index scheme is elongated,
then the whole index scheme is elongated. In the initial set of index schemes,
all schemes that were already interval schemes form a lower bound, while those
schemes and the new interval schemes generated by elongation form the upper
bound. The reason a lower bound may not be generated is that all indexing schemes
might need elongation.

@@

There is no direct translation from an interval scheme such as $\interv{-2}{1}{}
\times \interv{-1}{1}{}$ to region constants. The reason is region constants
\texttt{forward} and \texttt{backward} require upper and lower bounds to be 0
respectively and \texttt{centered} requires the upper bound to be the negation
of the lower bound. The solution is to use \Cref{lem:vector-union} to break this
interval scheme into two producing $\interv{-2}{0}{} \times \interv{-1}{1}{}$
and $\interv{0}{1}{} \times \interv{-1}{1}{}$.

@@

In the example, $\interv{-1}{1}{} \times \interv{0}{0}{}$ is decomposed into
$\interv{-1}{1}{} \times \interv{-\infty}{\infty}{}$ and
$\interv{-\infty}{\infty}{} \times \interv{0}{0}{}$. These are then mapped to
\texttt{\stencil{c}{1}{1}{} * \stencil{p}{2}{}{}}. The decomposition of
$\interv{0}{0}{} \times \interv{-1}{1}{}$ is similar. When combined we obtain the
following specification:
%
\begin{equation*}
  \texttt{stencil readOnce,} \;
  \stencil[s]{c}{1}{1}{} \texttt{*} \stencil[s]{p}{2}{}{} \; \texttt{+} \;
  \stencil[s]{p}{1}{}{}  \texttt{*} \stencil[s]{c}{2}{1}{}
\end{equation*}
